### 2021-05-10 proposal/proposal.Rmd
some more updates to proposal

# Research Goal

Typical clustering methods often work by assuming that each cluster is represented by some central point around which the data belonging to that cluster lie Such methods, including Lloyd's algorithm for $k$-means clustering \cite{1056489} and Gaussian Mixture Models \cite{doi:10.1198/016214502760047131}, involve iteratively updating the central points by averaging data belonging to each cluster and cluster memberships by comparing proximity to central points. The aim of our research is to explore clustering and community detection methods for which each cluster is represented not by a central point but by subspaces or manifolds. In addition, we aim to explore various applications of such methods and connect them to preexisting clustering and community detection problems.

@@

@@

It has been shown by \citet{jmlr-v28-wang13} that given sufficiently low noise in $X$ and sufficiently low affinity between pairs of subspaces as measured by the cosine of the angle between subspaces, SDP holds for a specific range of $\lambda$. This provides theoretical justification for applying SSC to Random Dot Product Graphs (RDPG) \cite{athreya2017statistical} and Generalized Random Dot Product Graphs (GRDPG) \cite{rubindelanchy2017statistical} for which the data in the latent space lie on subspaces.

\citeauthor{athreya2017statistical} showed that under mild conditions, if $(A_n, X_n) \sim \text{RDPG}(F, n)$, if $\hat{X}_n$ is the scaled ASE of $A_n$, for some sequence of orthogonal matrices $W_n$,

Similarly, \citeauthor{rubindelanchy2017statistical} showed that for $(A_n, X_n) \sim \text{GRDPG}_{p, q}(F, n)$,

Given sufficiently large sample size $n$, the scaled ASE of affinity matrix $A$ drawn from a RDPG or GRDPG will asymptotically approach the original latent positions $X$ with probability 1, up to a linear transformation (orthogonal transformation for the RDPG, a composition of an orthogonal transformation and scale transformation for the GRDPG). Thus if $X$ consists of points that lie on subspaces of $\mathbb{R}^d$, then both the scaled and unscaled ASE of $A \sim \text{RDPG}(X)$ or $A \sim \text{GRDPG}(X)$ will consist of points that lie near subspaces, with some noise that almost surely goes to $0$ as $n \to \infty$, motivating ASE followed by SSC as an asymptotically consistent method for community detection. The Popularity Adjusted Block Model (PABM) \cite{307cbeb9b1be48299388437423d94bf1} is a generative graph model with underlying communities such that each community lies on a subspace. \citet{noroozi2019estimation} showed that SSC is able to recover the subspaces and therefore perform community detection for the PABM given $P = X I_{p, q} X^\top$, the edge probability matrix, rather than $A$, the adjacency matrix. Combining the results of \citeauthor{rubindelanchy2017statistical} and \citeauthor{jmlr-v28-wang13}, it should be possible to recover the communities using $A$ as well.

@@

In addition to subspaces, \citet{trosset2020learning} showed that the ASE of a RDPG can be used to recover one-dimensional manifolds. Suppose $f : [0, 1] \mapsto \mathcal{X}$ such that $f$ is smooth and $\mathcal{X}$ represents a curve or one-dimensional manifold in $\mathbb{R}^d$. If $t_1, ..., t_n \stackrel{iid}{\sim} F$ such that $F$ has support $[0, 1]$, the latent positions are $x_i = f(t_i)$ with $y_i$ is its corresponding point in the scaled ASE, and $d_{\epsilon}(\cdot, \cdot)$ is the shortest path distance of an $\epsilon$-neighborhood graph. Under certain mild conditions, the shortest path distances of the $\epsilon$-neighborhood graph of the ASE approaches the arc lengths along $f$:

\begin{equation}
d_{\epsilon}(y_i, y_j) \stackrel{p}{\to} \int_{t_i}^{t_j} \sqrt{1  + (f'(t))^2} dt
\end{equation}

This suggests that if the latent positions of a RDPG consists of $K$ disjoint manifolds, the $\epsilon$-neighborhood graph will consist of $K$ disjoint subgraphs as $\epsilon \to 0$ and $n \to \infty$.

@@

@@

1. Define functions $f_1, ..., f_K$ such that 
$f_k : [0, 1] \mapsto \mathcal{X}$ and $f_i(t) \neq f_j(t)$ $\forall i \in [K]$ 
and $t \in [0, 1]$.
2. Sample labels
$z_1, ..., z_n \stackrel{iid}{\sim} Categorical(\pi_1, ..., \pi_K)$.
3. Sample $t_1, ..., t_n \stackrel{iid}{\sim} D$ where $D$ has support 
$[0, 1]$. 
4. Set latent positions $x_i = f_{z_i}(t_i)$ and 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.

>>>>

# Introduction

## Research Goal

Graph and network data have become increasingly more widespread in various fields including sociology, neuroscience, biostatistics, and computer science. This has resulted in certain challenges for researchers, as most traditional statistical and machine learning methods are incompatible with graphs and instead require the data to exist in a subset of Euclidean space. This includes clustering and community detection, which are the typical problems of interest when working with graphs. Common clustering methods often involve calculating some central or representative point around which the data belonging to that cluster lie (e.g., Lloyd's algorithm for $K$-means clustering \cite{1056489}, Gaussian Mixture Models \cite{doi:10.1198/016214502760047131}). Because these methods involve computing summary statistics within each cluster, they cannot be applied directly to graphs as community detection algorithms.

Some proposed methods for unifying graph community detection with traditional clustering techniques involve embedding the graph into Euclidean space, of which the most popular is Spectral Clustering \cite{DBLP:journals/corr/abs-0711-0189}. The Random Dot Product Graph (RDPG) \cite{10.1007/978-3-540-77004-6_11} and Generalized Random Dot Product Graph (GRDPG) \cite{rubindelanchy2017statistical} models take this further by explicitly constructing generative models such that latent positions in Euclidean space are used to induce graphs. A community detection method motivated by this would involve learning the latent positions given an observed graph and then learning the community labels given the latent positions.

The aim of our research is to develop consistent community detection techniques under the RDPG and GRDPG frameworks. First, we explore existing generative graph models with underlying community structures that can be inferred by connecting the generative models to the RDPG or GRDPG. Then we explore other latent structures or mixture distributions in the latent space that induce graphs for which consistent community detection is possible.

## Notation

Let $G = (V, E)$ be an undirected, unweighted graph with no self-loops with $n$ vertices. Denote $A \in \{0, 1\}^{n \times n}$ as the adjacency matrix for $G$ such that $A_{ij} = 1$ if there exists an edge between vertices $i$ and $j$ and $A_{ij} = 0$ otherwise. Because $G$ is symmetric and contains no self-loops, $A_{ij} = A_{ji}$ and $A_{ii} = 0$ for $i, j \in [n]$. We further restrict our analyses to independent Bernoulli graphs. Let $P \in [0, 1]^{n \times n}$ be a symmetric matrix of edge probabilities. Graph $G$ is sampled by $A_{ij} \indep P_{ij}$ for each $0 \leq i < j \leq n$ ($A_{ji} = A_{ij}$ and $A_{ii} = 0$). Denote $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}$ as the sample $x_1, ..., x_n \in \mathbb{R}^d$, and denote $z_1, ..., z_n \in [K]$ as their corresponding (hidden) labels.

@@

## Generative Graph Models and Community Detection

Generative models for symmetric Bernoulli graphs involve defining the edge probability matrix $P$ whose $ij$^th^ entry is the probability of an edge between vertices $i$ and $j$ for each $i, j \in [n]$. Community detection and recovery starts by restricting the generative model such that for each pair of vertices, the probability of an edge between the vertices is conditioned on the labels of the vertices. One such model is the Stochastic Block Model (SBM) \cite{doi:10.1080/0022250X.1971.9989788}: Given $K$ communities and each vertex belonging to one community, the SBM restricts $P$ to $K (K + 1) / 2$ unique entries such that $P_{ij} = B_{z_i z_j}$ where $B \in [0, 1]^{K \times K}$ with entries $B_{kl}$ denoting the edge probability of each vertex in community $k$ having an edge with each vertex in community $l$. The homogeneous SBM further restricts $P$ to two unique entries such that $P_{ij} = p$ if $z_i = z_j$ and $P_{ij} = q$ otherwise. Multiple generalizations of the SBM have been introduced since, including the Degree Corrected Block Model (DCBM) and the Popularity Adjusted Block Model (PABM). Like the SBM, these models involve edge probability matrix $P$ that is restricted to rank $d < n$ in part based on community labels.

The RDPG starts with points in latent space $X \in \mathcal{X} \subset \mathbb{R}^d$ such that $\forall x, y \in \mathcal{X}$, $x^\top y \in [0, 1]$. $P$ is then constructed as $P = X X^\top$, and graph $G$ with adjacency matrix $A$ is drawn from $P$. We provide a more formal definition of the RDPG and GRDPG below.

The structure of the RDPG and GRDPG provides an intuitive method for recovery of the latent positions via spectral embedding.

\citeauthor{athreya2017statistical} showed that under mild conditions, if $(A_n, X_n) \sim \text{RDPG}(F, n)$ and $\hat{X}_n$ is the scaled ASE of $A_n$, for some sequence of orthogonal matrices $W_n$,

Similarly, \citet{rubindelanchy2017statistical} showed that for $(A_n, X_n) \sim \text{GRDPG}_{p, q}(F, n)$,

It is straightforward to show that all Bernoulli graphs with positive semidefinite $P$ are special cases of the RDPG, which is a special case of the GRDPG, and all graphs generated by $P$ are special cases of the GRDPG. This includes the SBM, DCBM, and PABM.

\begin{example}[Connecting the SBM to the RDPG]
Let $G = (V, E)$ with adjacency matrix $A$ be sampled from the homogeneous SBM with two communities such that within-community edge probability $p$ and between-community edge probability $q$ where $p > q$. Let community 1 have $n_1$ vertices and community 2 have $n_2$ vertices such that $n_1 + n_2 = n$. Without loss of generality, organize $P$ and $A$ such that the $kl^{th}$ block represents edges between communities $k$ and $l$. Then $P = \begin{bmatrix} P^{(11)} & P^{(12)} \\ P^{(21)} & P^{(22)} \end{bmatrix}$ where each block is a constant value, e.g., $P^{(11)}_{ij} = p$. One RDPG representation of this SBM is:

$$X = \begin{bmatrix} 
\sqrt{p} & 0 \\
\vdots & \vdots \\
\sqrt{p} & 0 \\
\sqrt{r^2 / p} & \sqrt{q - r^2 / p} \\ 
\vdots & \vdots \\
\sqrt{r^2 / p} & \sqrt{q - r^2 / p}
\end{bmatrix}
\in \mathbb{R}^{n \times 2}$$

where the first $n_1$ rows are $\begin{bmatrix} \sqrt{p} & 0 \end{bmatrix}$ and the next $n_2$ rows are $\begin{bmatrix} \sqrt{r^2 / p} & \sqrt{q - r^2 / p} \end{bmatrix}$. Then it can be shown that

$$P = X X^\top$$
\end{example}

The ASE of the assortative SBM consists of points in $\mathbb{R}^K$ that lie near one of $K$ centers, depending on the community label, leading to ASE followed by $K$-means clustering (or similar, e.g., GMM) as a consistent community detection algorithm \cite{lyzinski2014}. A similar result can be shown for the DCBM and PABM but with different structures in the ASE.

Given sufficiently large sample size $n$, the scaled ASE of affinity matrix $A$ drawn from a RDPG or GRDPG will asymptotically approach the original latent positions $X$ with probability 1, up to a linear transformation (orthogonal transformation for the RDPG, a composition of an orthogonal and scale transformations for the GRDPG). Thus if $X$ consists of points that lie on subspaces of $\mathbb{R}^d$, then both the scaled and unscaled ASE of $A \sim \text{RDPG}(X)$ or $A \sim \text{GRDPG}(X)$ will consist of points that lie near subspaces, with some noise that almost surely goes to $0$ as $n \to \infty$, motivating ASE followed by SSC as an asymptotically consistent method for community detection. The Popularity Adjusted Block Model (PABM) \cite{307cbeb9b1be48299388437423d94bf1} is a generative graph model with underlying communities such that each community lies on a subspace. \citet{noroozi2019estimation} showed that SSC is able to recover the subspaces and therefore perform community detection for the PABM given $P = X I_{p, q} X^\top$, the edge probability matrix, rather than $A$, the adjacency matrix. Combining the results of \citeauthor{rubindelanchy2017statistical} and \citeauthor{jmlr-v28-wang13}, it should be possible to recover the communities using $A$ as well.

@@

@@

\citet{trosset2020learning} showed that the ASE of a RDPG can be used to recover one-dimensional manifolds. Suppose $f : [0, 1] \mapsto \mathcal{X}$ such that $f$ is smooth and $\mathcal{X}$ represents a curve or one-dimensional manifold in $\mathbb{R}^d$. If $t_1, ..., t_n \stackrel{iid}{\sim} F$ such that $F$ has support $[0, 1]$, the latent positions are $x_i = f(t_i)$ with $y_i$ is its corresponding point in the scaled ASE, and $d_{\epsilon}(\cdot, \cdot)$ is the shortest path distance of an $\epsilon$-neighborhood graph. Under certain mild conditions, the shortest path distances of the $\epsilon$-neighborhood graph of the ASE approaches the arc lengths along $f$:

\begin{equation}
d_{\epsilon}(y_i, y_j) \stackrel{p}{\to} \int_{t_i}^{t_j} \sqrt{\sum_r^d \Big( \frac{d f_r}{d t} \Big)^2} dt
\end{equation}

\citet{athreya2020estimation} extended this further by generating a RDPG from a mixture of distributions on a curve. In their example, points were sampled from a mixture of two Beta distributions on the Hardy-Weinberg curve to construct the latent positions of a RDPG, with the goal of recovering the hidden mixture distribution from an observed graph.

@@

## Connecting Existing Generative Graph Models to the RDPG and GRDPG

### Subspace Clustering

@@

1. Define functions $f_1, ..., f_K$ such that 
$f_k : [0, 1] \mapsto \mathcal{X}$ and $f_k(t) \neq f_l(t)$ $\forall k, l \in [K]$ 
and $t \in [0, 1]$.
2. Sample labels
$z_1, ..., z_n \stackrel{iid}{\sim} Categorical(\pi_1, ..., \pi_K)$.
3. Sample $t_1, ..., t_n \stackrel{iid}{\sim} D$ where $D$ has support 
$[0, 1]$. 
4. Set latent positions $x_i = f_{z_i}(t_i)$ and 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.

### 2021-05-10 proposal/proposal.tex
some more updates to proposal

%\begin{titlepage}
\begin{center}
\LARGE{\textbf{Community Detection on Subspaces and Manifolds}}\\
\vspace*{2\baselineskip}
\normalsize{A dissertation proposal submitted in partial satisfaction of the requirements for the degree of \\}
Doctor of Philosophy \\
in \\
Statistical Science \\
\vspace*{2\baselineskip}
\Large{John Koo}\\
\vspace*{3\baselineskip}
\Large{\textbf{Research Committee Members}}\\
Dr. Michael Trosset \\
Dr. Minh Tang \\
Dr. Julia Fukuyama \\
Dr. Roni Khardon \\
Dr. Fangzheng Xie \\
\vspace*{3\baselineskip}
Date TBA \\
\vspace*{1\baselineskip}
Department of Statistics \\
Indiana University \\
Bloomington, Indiana \\
\end{center}
% \end{titlepage}

@@

\hypertarget{research-goal}{%
\section{Research Goal}\label{research-goal}}

Typical clustering methods often work by assuming that each cluster is
represented by some central point around which the data belonging to
that cluster lie Such methods, including Lloyd's algorithm for
\(k\)-means clustering \cite{1056489} and Gaussian Mixture Models
\cite{doi:10.1198/016214502760047131}, involve iteratively updating the
central points by averaging data belonging to each cluster and cluster
memberships by comparing proximity to central points. The aim of our
research is to explore clustering and community detection methods for
which each cluster is represented not by a central point but by
subspaces or manifolds. In addition, we aim to explore various
applications of such methods and connect them to preexisting clustering
and community detection problems.

@@

@@

It has been shown by \citet{jmlr-v28-wang13} that given sufficiently low
noise in \(X\) and sufficiently low affinity between pairs of subspaces
as measured by the cosine of the angle between subspaces, SDP holds for
a specific range of \(\lambda\). This provides theoretical justification
for applying SSC to Random Dot Product Graphs (RDPG)
\cite{athreya2017statistical} and Generalized Random Dot Product Graphs
(GRDPG) \cite{rubindelanchy2017statistical} for which the data in the
latent space lie on subspaces.

\citeauthor{athreya2017statistical} showed that under mild conditions,
if \((A_n, X_n) \sim \text{RDPG}(F, n)\), if \(\hat{X}_n\) is the scaled
ASE of \(A_n\), for some sequence of orthogonal matrices \(W_n\),

Similarly, \citeauthor{rubindelanchy2017statistical} showed that for
\((A_n, X_n) \sim \text{GRDPG}_{p, q}(F, n)\),

Given sufficiently large sample size \(n\), the scaled ASE of affinity
matrix \(A\) drawn from a RDPG or GRDPG will asymptotically approach the
original latent positions \(X\) with probability 1, up to a linear
transformation (orthogonal transformation for the RDPG, a composition of
an orthogonal transformation and scale transformation for the GRDPG).
Thus if \(X\) consists of points that lie on subspaces of
\(\mathbb{R}^d\), then both the scaled and unscaled ASE of
\(A \sim \text{RDPG}(X)\) or \(A \sim \text{GRDPG}(X)\) will consist of
points that lie near subspaces, with some noise that almost surely goes
to \(0\) as \(n \to \infty\), motivating ASE followed by SSC as an
asymptotically consistent method for community detection. The Popularity
Adjusted Block Model (PABM) \cite{307cbeb9b1be48299388437423d94bf1} is a
generative graph model with underlying communities such that each
community lies on a subspace. \citet{noroozi2019estimation} showed that
SSC is able to recover the subspaces and therefore perform community
detection for the PABM given \(P = X I_{p, q} X^\top\), the edge
probability matrix, rather than \(A\), the adjacency matrix. Combining
the results of \citeauthor{rubindelanchy2017statistical} and
\citeauthor{jmlr-v28-wang13}, it should be possible to recover the
communities using \(A\) as well.

@@

In addition to subspaces, \citet{trosset2020learning} showed that the
ASE of a RDPG can be used to recover one-dimensional manifolds. Suppose
\(f : [0, 1] \mapsto \mathcal{X}\) such that \(f\) is smooth and
\(\mathcal{X}\) represents a curve or one-dimensional manifold in
\(\mathbb{R}^d\). If \(t_1, ..., t_n \stackrel{iid}{\sim} F\) such that
\(F\) has support \([0, 1]\), the latent positions are \(x_i = f(t_i)\)
with \(y_i\) is its corresponding point in the scaled ASE, and
\(d_{\epsilon}(\cdot, \cdot)\) is the shortest path distance of an
\(\epsilon\)-neighborhood graph. Under certain mild conditions, the
shortest path distances of the \(\epsilon\)-neighborhood graph of the
ASE approaches the arc lengths along \(f\):

\begin{equation}
d_{\epsilon}(y_i, y_j) \stackrel{p}{\to} \int_{t_i}^{t_j} \sqrt{1  + (f'(t))^2} dt
\end{equation}

This suggests that if the latent positions of a RDPG consists of \(K\)
disjoint manifolds, the \(\epsilon\)-neighborhood graph will consist of
\(K\) disjoint subgraphs as \(\epsilon \to 0\) and \(n \to \infty\).

@@

\hypertarget{subspace-clustering-1}{%
\subsection{Subspace Clustering}\label{subspace-clustering-1}}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Define functions \(f_1, ..., f_K\) such that
  \(f_k : [0, 1] \mapsto \mathcal{X}\) and \(f_i(t) \neq f_j(t)\)
  \(\forall i \in [K]\) and \(t \in [0, 1]\).
\item
  Sample labels
  \(z_1, ..., z_n \stackrel{iid}{\sim} Categorical(\pi_1, ..., \pi_K)\).
\item
  Sample \(t_1, ..., t_n \stackrel{iid}{\sim} D\) where \(D\) has
  support \([0, 1]\).
\item
  Set latent positions \(x_i = f_{z_i}(t_i)\) and
  \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\).
\end{enumerate}

>>>>

%\begin{titlepage}
\begin{center}
\LARGE{\textbf{Community Detection Methods for Random Dot Product Graphs and Generalized Random Dot Product Graphs}}\\
\vspace*{2\baselineskip}
\normalsize{A dissertation proposal submitted in partial satisfaction of the requirements for the degree of \\}
Doctor of Philosophy \\
in \\
Statistical Science \\
\vspace*{2\baselineskip}
\Large{John Koo}\\
\vspace*{3\baselineskip}
\Large{\textbf{Research Committee Members}}\\
Dr. Michael Trosset \\
Dr. Minh Tang \\
Dr. Julia Fukuyama \\
Dr. Roni Khardon \\
Dr. Fangzheng Xie \\
\vspace*{3\baselineskip}
Date TBA \\
\vspace*{1\baselineskip}
Department of Statistics \\
Indiana University \\
Bloomington, Indiana \\
\end{center}
% \end{titlepage}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\hypertarget{research-goal}{%
\subsection{Research Goal}\label{research-goal}}

Graph and network data have become increasingly more widespread in
various fields including sociology, neuroscience, biostatistics, and
computer science. This has resulted in certain challenges for
researchers, as most traditional statistical and machine learning
methods are incompatible with graphs and instead require the data to
exist in a subset of Euclidean space. This includes clustering and
community detection, which are the typical problems of interest when
working with graphs. Common clustering methods often involve calculating
some central or representative point around which the data belonging to
that cluster lie (e.g., Lloyd's algorithm for \(K\)-means clustering
\cite{1056489}, Gaussian Mixture Models
\cite{doi:10.1198/016214502760047131}). Because these methods involve
computing summary statistics within each cluster, they cannot be applied
directly to graphs as community detection algorithms.

Some proposed methods for unifying graph community detection with
traditional clustering techniques involve embedding the graph into
Euclidean space, of which the most popular is Spectral Clustering
\cite{DBLP:journals/corr/abs-0711-0189}. The Random Dot Product Graph
(RDPG) \cite{10.1007/978-3-540-77004-6_11} and Generalized Random Dot
Product Graph (GRDPG) \cite{rubindelanchy2017statistical} models take
this further by explicitly constructing generative models such that
latent positions in Euclidean space are used to induce graphs. A
community detection method motivated by this would involve learning the
latent positions given an observed graph and then learning the community
labels given the latent positions.

The aim of our research is to develop consistent community detection
techniques under the RDPG and GRDPG frameworks. First, we explore
existing generative graph models with underlying community structures
that can be inferred by connecting the generative models to the RDPG or
GRDPG. Then we explore other latent structures or mixture distributions
in the latent space that induce graphs for which consistent community
detection is possible.

\hypertarget{notation}{%
\subsection{Notation}\label{notation}}

Let \(G = (V, E)\) be an undirected, unweighted graph with no self-loops
with \(n\) vertices. Denote \(A \in \{0, 1\}^{n \times n}\) as the
adjacency matrix for \(G\) such that \(A_{ij} = 1\) if there exists an
edge between vertices \(i\) and \(j\) and \(A_{ij} = 0\) otherwise.
Because \(G\) is symmetric and contains no self-loops,
\(A_{ij} = A_{ji}\) and \(A_{ii} = 0\) for \(i, j \in [n]\). We further
restrict our analyses to independent Bernoulli graphs. Let
\(P \in [0, 1]^{n \times n}\) be a symmetric matrix of edge
probabilities. Graph \(G\) is sampled by
\(A_{ij} \stackrel{\text{indep}}{\sim}P_{ij}\) for each
\(0 \leq i < j \leq n\) (\(A_{ji} = A_{ij}\) and \(A_{ii} = 0\)). Denote
\(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}\)
as the sample \(x_1, ..., x_n \in \mathbb{R}^d\), and denote
\(z_1, ..., z_n \in [K]\) as their corresponding (hidden) labels.

@@

\hypertarget{generative-graph-models-and-community-detection}{%
\subsection{Generative Graph Models and Community
Detection}\label{generative-graph-models-and-community-detection}}

Generative models for symmetric Bernoulli graphs involve defining the
edge probability matrix \(P\) whose \(ij\)\textsuperscript{th} entry is
the probability of an edge between vertices \(i\) and \(j\) for each
\(i, j \in [n]\). Community detection and recovery starts by restricting
the generative model such that for each pair of vertices, the
probability of an edge between the vertices is conditioned on the labels
of the vertices. One such model is the Stochastic Block Model (SBM)
\cite{doi:10.1080/0022250X.1971.9989788}: Given \(K\) communities and
each vertex belonging to one community, the SBM restricts \(P\) to
\(K (K + 1) / 2\) unique entries such that \(P_{ij} = B_{z_i z_j}\)
where \(B \in [0, 1]^{K \times K}\) with entries \(B_{kl}\) denoting the
edge probability of each vertex in community \(k\) having an edge with
each vertex in community \(l\). The homogeneous SBM further restricts
\(P\) to two unique entries such that \(P_{ij} = p\) if \(z_i = z_j\)
and \(P_{ij} = q\) otherwise. Multiple generalizations of the SBM have
been introduced since, including the Degree Corrected Block Model (DCBM)
and the Popularity Adjusted Block Model (PABM). Like the SBM, these
models involve edge probability matrix \(P\) that is restricted to rank
\(d < n\) in part based on community labels.

The RDPG starts with points in latent space
\(X \in \mathcal{X} \subset \mathbb{R}^d\) such that
\(\forall x, y \in \mathcal{X}\), \(x^\top y \in [0, 1]\). \(P\) is then
constructed as \(P = X X^\top\), and graph \(G\) with adjacency matrix
\(A\) is drawn from \(P\). We provide a more formal definition of the
RDPG and GRDPG below.

The structure of the RDPG and GRDPG provides an intuitive method for
recovery of the latent positions via spectral embedding.

\citeauthor{athreya2017statistical} showed that under mild conditions,
if \((A_n, X_n) \sim \text{RDPG}(F, n)\) and \(\hat{X}_n\) is the scaled
ASE of \(A_n\), for some sequence of orthogonal matrices \(W_n\),

Similarly, \citet{rubindelanchy2017statistical} showed that for
\((A_n, X_n) \sim \text{GRDPG}_{p, q}(F, n)\),

It is straightforward to show that all Bernoulli graphs with positive
semidefinite \(P\) are special cases of the RDPG, which is a special
case of the GRDPG, and all graphs generated by \(P\) are special cases
of the GRDPG. This includes the SBM, DCBM, and PABM.

\begin{example}[Connecting the SBM to the RDPG]
Let $G = (V, E)$ with adjacency matrix $A$ be sampled from the homogeneous SBM with two communities such that within-community edge probability $p$ and between-community edge probability $q$ where $p > q$. Let community 1 have $n_1$ vertices and community 2 have $n_2$ vertices such that $n_1 + n_2 = n$. Without loss of generality, organize $P$ and $A$ such that the $kl^{th}$ block represents edges between communities $k$ and $l$. Then $P = \begin{bmatrix} P^{(11)} & P^{(12)} \\ P^{(21)} & P^{(22)} \end{bmatrix}$ where each block is a constant value, e.g., $P^{(11)}_{ij} = p$. One RDPG representation of this SBM is:

$$X = \begin{bmatrix} 
\sqrt{p} & 0 \\
\vdots & \vdots \\
\sqrt{p} & 0 \\
\sqrt{r^2 / p} & \sqrt{q - r^2 / p} \\ 
\vdots & \vdots \\
\sqrt{r^2 / p} & \sqrt{q - r^2 / p}
\end{bmatrix}
\in \mathbb{R}^{n \times 2}$$

where the first $n_1$ rows are $\begin{bmatrix} \sqrt{p} & 0 \end{bmatrix}$ and the next $n_2$ rows are $\begin{bmatrix} \sqrt{r^2 / p} & \sqrt{q - r^2 / p} \end{bmatrix}$. Then it can be shown that

$$P = X X^\top$$
\end{example}

The ASE of the assortative SBM consists of points in \(\mathbb{R}^K\)
that lie near one of \(K\) centers, depending on the community label,
leading to ASE followed by \(K\)-means clustering (or similar, e.g.,
GMM) as a consistent community detection algorithm \cite{lyzinski2014}.
A similar result can be shown for the DCBM and PABM but with different
structures in the ASE.

Given sufficiently large sample size \(n\), the scaled ASE of affinity
matrix \(A\) drawn from a RDPG or GRDPG will asymptotically approach the
original latent positions \(X\) with probability 1, up to a linear
transformation (orthogonal transformation for the RDPG, a composition of
an orthogonal and scale transformations for the GRDPG). Thus if \(X\)
consists of points that lie on subspaces of \(\mathbb{R}^d\), then both
the scaled and unscaled ASE of \(A \sim \text{RDPG}(X)\) or
\(A \sim \text{GRDPG}(X)\) will consist of points that lie near
subspaces, with some noise that almost surely goes to \(0\) as
\(n \to \infty\), motivating ASE followed by SSC as an asymptotically
consistent method for community detection. The Popularity Adjusted Block
Model (PABM) \cite{307cbeb9b1be48299388437423d94bf1} is a generative
graph model with underlying communities such that each community lies on
a subspace. \citet{noroozi2019estimation} showed that SSC is able to
recover the subspaces and therefore perform community detection for the
PABM given \(P = X I_{p, q} X^\top\), the edge probability matrix,
rather than \(A\), the adjacency matrix. Combining the results of
\citeauthor{rubindelanchy2017statistical} and
\citeauthor{jmlr-v28-wang13}, it should be possible to recover the
communities using \(A\) as well.

@@

@@

\citet{trosset2020learning} showed that the ASE of a RDPG can be used to
recover one-dimensional manifolds. Suppose
\(f : [0, 1] \mapsto \mathcal{X}\) such that \(f\) is smooth and
\(\mathcal{X}\) represents a curve or one-dimensional manifold in
\(\mathbb{R}^d\). If \(t_1, ..., t_n \stackrel{iid}{\sim} F\) such that
\(F\) has support \([0, 1]\), the latent positions are \(x_i = f(t_i)\)
with \(y_i\) is its corresponding point in the scaled ASE, and
\(d_{\epsilon}(\cdot, \cdot)\) is the shortest path distance of an
\(\epsilon\)-neighborhood graph. Under certain mild conditions, the
shortest path distances of the \(\epsilon\)-neighborhood graph of the
ASE approaches the arc lengths along \(f\):

\begin{equation}
d_{\epsilon}(y_i, y_j) \stackrel{p}{\to} \int_{t_i}^{t_j} \sqrt{\sum_r^d \Big( \frac{d f_r}{d t} \Big)^2} dt
\end{equation}

\citet{athreya2020estimation} extended this further by generating a RDPG
from a mixture of distributions on a curve. In their example, points
were sampled from a mixture of two Beta distributions on the
Hardy-Weinberg curve to construct the latent positions of a RDPG, with
the goal of recovering the hidden mixture distribution from an observed
graph.

@@

\hypertarget{connecting-existing-generative-graph-models-to-the-rdpg-and-grdpg}{%
\subsection{Connecting Existing Generative Graph Models to the RDPG and
GRDPG}\label{connecting-existing-generative-graph-models-to-the-rdpg-and-grdpg}}

\hypertarget{subspace-clustering-1}{%
\subsubsection{Subspace Clustering}\label{subspace-clustering-1}}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Define functions \(f_1, ..., f_K\) such that
  \(f_k : [0, 1] \mapsto \mathcal{X}\) and \(f_k(t) \neq f_l(t)\)
  \(\forall k, l \in [K]\) and \(t \in [0, 1]\).
\item
  Sample labels
  \(z_1, ..., z_n \stackrel{iid}{\sim} Categorical(\pi_1, ..., \pi_K)\).
\item
  Sample \(t_1, ..., t_n \stackrel{iid}{\sim} D\) where \(D\) has
  support \([0, 1]\).
\item
  Set latent positions \(x_i = f_{z_i}(t_i)\) and
  \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\).
\end{enumerate}
